Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-10-11 prop_letter.tex
prop_letter.tex \section{Born} and \section{Gleason} Updating...

On the one hand, $\iota_{1}$ defined in Eq.~(\ref{eq:iota}) satisfies
Eqs.~(\ref{eq:iota-constraints}) and Eq.~(\ref{eq:iota-add})~\citep{HOSTunpublished}
so that every function~$\iota_{1}\circ\mu$ on a three-dimensional
Hilbert space is a QIVPM. On the other hand, $\iota_{1}\circ\mu_{\rho_{1}}^{\mathrm{\mathrm{B}}}$
is not a QIVPM on four-dimensional Hilbert space. Therefore, we need
to strengthen Eqs.~(\ref{eq:iota-constraints}) and Eq.~(\ref{eq:iota-add})
to guarantee $\iota\circ\mu$ is always a QIVPM.

@@

In order to find a function~$\iota$ satisfying Eq.~(\ref{eq:iota-Inclusion}),
we decompose it into two functions: its left-end and right-end~$\left[\iotal\left(x\right),\iotar\left(x\right)\right]=\iota\left(x\right)$.
By Eq.~(\ref{eq:iota-Inclusion}), the left-end function~$\iotal:\left[0,1\right]\rightarrow\left[0,1\right]$
is Wright-convex, i.e., $\iotal\left(x_{0}+x_{1}+x_{2}\right)+\iotal\left(x_{2}\right)\ge\iotal\left(x_{0}+x_{2}\right)+\iotal\left(x_{1}+x_{2}\right)$,
and the right-end function~$\iotar:\left[0,1\right]\rightarrow\left[0,1\right]$
is Wright-concave, i.e., $\iotar\left(x_{0}+x_{1}+x_{2}\right)+\iotar\left(x_{2}\right)\le\iotar\left(x_{0}+x_{2}\right)+\iotar\left(x_{1}+x_{2}\right)$~\citep{Wright1954,RobertsVarberg1973,PecaricTong1992,Brzdek_2014}.
It is well-known that a Wright-convex function is midpoint-convex
or Jensen-convex~\citep{Klee_1956,Kenyon_1956,RobertsVarberg1973,PecaricTong1992},
i.e., we have
\begin{equation}
\frac{\iotal\left(x_{0}\right)+\iotal\left(x_{1}\right)}{2}\ge\iotal\left(\frac{x_{0}+x_{1}}{2}\right)\,.
\end{equation}
Together with the fact that $\iotal$ maps to a bounded interval~$\left[0,1\right]$,
the left-end function~$\iotal$ must be continuous on the unit open
interval~$\left(0,1\right)$~\citep{Jensen_1906,RobertsVarberg1973}
which implies the number of intervals that $\iota$ maps to must be
uncountable.

\begin{thm}For any function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
described in Thm.~\ref{thm:iota-statements}, either $\mathscr{I}=\left\{ \necess,\imposs,\unknown\right\} $
or $\mathscr{I}$ contains uncountably many intervals.\end{thm}

\yutsung{Another idea might be fixing a quantum probability measure~$\mu$,
whether there is a length of intervals, such that if $\iota$ mapping
to a interval tighter than the given length, then $\iota\circ\mu$
is always a QIVPM... I don't know the answer of this question.}

@@

\noindent However, if we consider general QIVPMs mapping to an \emph{arbitrary}
set of intervals, there is a QIVPM that no state can be consistent
with it.

@@

To prove the theorem, we first need to construct a QIVPM, and then
verify that no states consistent with it on all projectors. Given
an orthonormal basis $\left\{ \ket{0},\ket{1},\ket{2}\right\} $,
let $\mathscr{I}_{0}=\left\{ \necess,\imposs,\unknown\right\} $,
$\ket{\ps}=\left(\ket{0}+\ket{1}\right)/\sqrt{2}$, and $\ket{\ps'}=\left(\ket{0}+\ket{2}\right)/\sqrt{2}$.
The function~$\bar{\mu}:\events\rightarrow\mathscr{I}_{0}$ defined
in Table~\ref{tab:non-Born-QIVPM} can be verified as a QIVPM straightforwardly~\citep{HOSTunpublished}.
\begin{table}
\caption{\label{tab:non-Born-QIVPM}Define a QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}_{0}$
on the three-dimensional Hilbert space. The projectors mapping to
the same interval are listed in the column labeled by $P$ and separated
by commas.}

@@

In other words, if the measurement intervals are too imprecise, the
results we get might contradict to each other, and exclude any possible
state from being consistent with the measurement results. In contrast
with the QIVPM defined in Table~\ref{tab:non-Born-QIVPM}, if a QIVPM~$\iota\circ\mu_{\rho}^{\mathrm{B}}$
is induced by a conventional quantum probability measure~$\mu_{\rho}^{\mathrm{B}}$
and a function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$ defined
in Sec.~\ref{sec:Born}, there is always a state consistent with
it because $\rho\in\coreBorn\left(\iota\circ\mu_{\rho}^{\mathrm{B}},\events\right)$.
Moreover, with respect to a notion of distance between two states~$\rho$
and $\rho'$~\citep{544199}, 
\begin{equation}
\left\Vert \rho-\rho'\right\Vert =\max_{\ip{\psi}{\psi}=1}\left|\melem{\psi}{\rho-\rho'}{\psi}\right|\,,\label{eq:norm}
\end{equation}
the collection of states consistent with $\iota\circ\mu_{\rho}^{\mathrm{B}}$
gets closer and closer to $\rho$ as the intervals get narrower and
narrower.

\begin{thm}[A finite-precision extension of weak Gleason's theorem
for Born QIVPMs]\label{thm:Finite-precision-Gleason}Consider a three-dimensional
Hilbert space, and an interval-valued function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
with the maximum length of intervals in $\mathscr{I}$ is $\delta$.
Given any state~$\rho$, if a state $\rho'\in\coreBorn\left(\iota\circ\mu_{\rho}^{\mathrm{B}},\events\right)$,
then their distance $\left\Vert \rho-\rho'\right\Vert \le\delta$.\end{thm}

\noindent Given a state~$\rho'$ consistent with $\iota\circ\mu_{\rho}^{\mathrm{B}}$,
we have $\mu_{\rho'}^{\mathrm{B}}\left(\proj{\psi}\right)\in\iota\left(\mu_{\rho}^{\mathrm{B}}\left(\proj{\psi}\right)\right)$
for any 1-dimensional projector~$\proj{\psi}$. Since the maximum
length of the intervals in $\mathscr{I}$ is $\delta$, it is also
the upper bound of the difference $\left|\mu_{\rho'}^{\mathrm{B}}\left(\proj{\psi}\right)-\mu_{\rho}^{\mathrm{B}}\left(\proj{\psi}\right)\right|=\left|\melem{\psi}{\rho-\rho'}{\psi}\right|$.
This implies $\left\Vert \rho-\rho'\right\Vert \le\delta$ by definition.

\yutsung{Following are the discussion and personal notes about our
norm and other notation of norm and distance.}

Nielsen and Chuang~\citep{544199} defined their norm~(\ref{eq:norm})
in Appendix~6. While their Appendix~6 followed Simon~\citep{Simon2005}
and Uhlmann~\citep{Uhlmann1977} to prove Lieb's theorem~\citep{Lieb_1973,544199,Simon2005},
Simon~\citep{Simon2005} and Uhlmann~\citep{Uhlmann1977} didn't
explicitly define the norm as in Eq.~(\ref{eq:norm}) although Simon
used the same norm thoroughly in his book~\citep{Simon2005}. There
are two possibilities:
\begin{enumerate}
\item Simon really used Eq.~(\ref{eq:norm}) to define his norm.\\
If this is true, there must be some books cited by Simon, and defined
the norm in the same way.
\item The norm Simon used is different from Eq.~(\ref{eq:norm}), and Nielsen
and Chuang defined this norm from nowhere in order to write a proof
of Lieb's theorem.\\
If this is true, Nielsen and Chuang might be the \emph{only} literature
which used this norm. We might have no other literature to check the
properties of this norm and other distances, and need to prove the
properties of this norm by ourselves.
\end{enumerate}
In order to understand this norm and compare with others, the following
are useful facts. Let $\sigma\left(\rho-\rho'\right)$ denote the
set of eigenvalues of $\rho-\rho'$. Since $\rho$ and $\rho'$ are
states, they are Hermitian~\citep{544199}, and so does $\rho-\rho'$.
In this situation~\citep{544199}, 
\begin{equation}
\left\Vert \rho-\rho'\right\Vert =\max_{\lambda\in\sigma\left(\rho-\rho'\right)}\left|\lambda\right|
\end{equation}
which is just its spectral radius or spectral norm~\citep{GolubVanLoan1996,Foucart2012}
\begin{equation}
\max_{\left\Vert \ket{\psi}\right\Vert =1}\left\Vert \left(\rho-\rho'\right)\ket{\psi}\right\Vert \,.
\end{equation}
For their trace distance~\citep{544199}
\begin{equation}
D\left(\rho,\rho'\right)=\frac{1}{2}\Tr\left[\sqrt{\left(\rho-\rho'\right)^{2}}\right]=\frac{1}{2}\sum_{\lambda\in\sigma\left(\rho-\rho'\right)}\left|\lambda\right|\,,
\end{equation}
since it can also be represented as the eigenvalue of $\rho-\rho'$.
It should not be hard to estimate their trace distance based on their
norm.

\yutsung{End of discussion.}

>>>>

On the one hand, $\iota_{1}$ defined in Eq.~(\ref{eq:iota}) satisfies
Eqs.~(\ref{eq:iota-constraints}) and Eq.~(\ref{eq:iota-add})~\citep{HOSTunpublished}
so that every function~$\iota_{1}\circ\mu$ on a three-dimensional
Hilbert space is a QIVPM. On the other hand, $\iota_{1}\circ\mu_{\rho_{1}}^{\mathrm{\mathrm{B}}}$
is not a QIVPM on four-dimensional Hilbert space. Hence, to guarantee
$\iota\circ\mu$ is always a QIVPM, we need to strengthen Eq.~(\ref{eq:iota-add})
and consider the following conditions.

@@

In order to find a function~$\iota$ satisfying Eq.~(\ref{eq:iota-Inclusion}),
we decompose it into two functions: its left-end and right-end~$\left[\iotal\left(x\right),\iotar\left(x\right)\right]=\iota\left(x\right)$.
By Eq.~(\ref{eq:iota-Inclusion}), the left-end function~$\iotal:\left[0,1\right]\rightarrow\left[0,1\right]$
is Wright-convex, i.e., $\iotal\left(x_{0}+x_{1}+x_{2}\right)+\iotal\left(x_{2}\right)\ge\iotal\left(x_{0}+x_{2}\right)+\iotal\left(x_{1}+x_{2}\right)$,
and the right-end function~$\iotar:\left[0,1\right]\rightarrow\left[0,1\right]$
is Wright-concave, i.e., $\iotar\left(x_{0}+x_{1}+x_{2}\right)+\iotar\left(x_{2}\right)\le\iotar\left(x_{0}+x_{2}\right)+\iotar\left(x_{1}+x_{2}\right)$~\citep{Wright1954,RobertsVarberg1973,PecaricTong1992,Brzdek_2014}.
It is well-known that a Wright-convex function is midpoint-convex
or Jensen-convex~\citep{Klee_1956,Kenyon_1956,RobertsVarberg1973,PecaricTong1992},
i.e., we have
\begin{equation}
\frac{\iotal\left(x_{0}\right)+\iotal\left(x_{1}\right)}{2}\ge\iotal\left(\frac{x_{0}+x_{1}}{2}\right)\,.
\end{equation}
Together with the fact that $\iotal$ maps to a bounded interval~$\left[0,1\right]$,
the left-end function~$\iotal$ must be continuous on the unit open
interval~$\left(0,1\right)$~\citep{Jensen_1906,RobertsVarberg1973}.
Therefore, either $\iota$ maps every number in the same interval,
or the number of intervals that $\iota$ maps to must be uncountable.

\begin{thm}\label{thm:convex-uncountable}Let $\mathscr{I}_{0}=\left\{ \necess,\imposs,\unknown\right\} $.
For any function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
described in Thm.~\ref{thm:iota-statements}, either $\mathscr{I}$
is $\mathscr{I}_{0}$ or $\mathscr{I}$ contains uncountably many
intervals.\end{thm}

\yutsung{Another idea might be fixing a quantum probability measure~$\mu$,
whether there is a length of intervals, such that if $\iota$ maps
to intervals tighter than the given length, then $\iota\circ\mu$
is always a QIVPM... I don't know the answer of this question, and
it might not be very easy...}

@@

However, if we consider general QIVPMs mapping to an \emph{arbitrary}
set of intervals, there is a QIVPM that no state can be consistent
with it.

@@

To prove the theorem, we first need to construct a QIVPM, and then
verify that no states consistent with it on all projectors. Given
an orthonormal basis $\left\{ \ket{0},\ket{1},\ket{2}\right\} $,
let $\ket{\ps}=\left(\ket{0}+\ket{1}\right)/\sqrt{2}$ and $\ket{\ps'}=\left(\ket{0}+\ket{2}\right)/\sqrt{2}$.
The function~$\bar{\mu}:\events\rightarrow\mathscr{I}_{0}$ defined
in Table~\ref{tab:non-Born-QIVPM} can be verified as a QIVPM straightforwardly~\citep{HOSTunpublished},
where $\mathscr{I}_{0}$ is defined in Thm.~\ref{thm:convex-uncountable}.
\begin{table}
\caption{\label{tab:non-Born-QIVPM}Define a QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}_{0}$
on the three-dimensional Hilbert space. The projectors mapping to
the same interval are listed in the column labeled by $P$ and separated
by commas.}

@@

In other words, if the measurement intervals are too imprecise, the
results we get might contradict to each other, and exclude any possible
state from being consistent with the measurement results. In contrast
with the QIVPM defined in Table~\ref{tab:non-Born-QIVPM}, if a QIVPM~$\iota\circ\mu_{\rho}^{\mathrm{B}}$
is induced by a conventional quantum probability measure~$\mu_{\rho}^{\mathrm{B}}$
and a function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$ considered
in Sec.~\ref{sec:Born}, there is always a state consistent with
it because $\rho\in\coreBorn\left(\iota\circ\mu_{\rho}^{\mathrm{B}},\events\right)$.
Moreover, the collection of states consistent with $\iota\circ\mu_{\rho}^{\mathrm{B}}$
gets closer and closer to $\rho$ as the intervals get narrower and
narrower. To state this result, we introduce a few technical notions.
First, the norm of a pure state is defined as usual $||\psi||=\sqrt{\ip{\psi}{\psi}}$.
Then, the norm of the difference of two states~$\rho$ and $\rho'$
is just the usual operator norm
\begin{equation}
\left\Vert \rho-\rho'\right\Vert =\max_{\left\Vert \psi\right\Vert =1}\left\Vert \left(\rho-\rho'\right)\ket{\psi}\right\Vert \,,
\end{equation}
which is also commonly known as $L^{2}$ norm or spectral norm~\citep{GolubVanLoan1996,Foucart2012}.

\begin{thm}[A finite-precision extension of weak Gleason's theorem
for Born QIVPMs]\label{thm:Finite-precision-Gleason}Consider a three-dimensional
Hilbert space, and an interval-valued function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
with the maximum length of intervals in $\mathscr{I}$ is $\delta$.
Given any state~$\rho$, if a state $\rho'\in\coreBorn\left(\iota\circ\mu_{\rho}^{\mathrm{B}},\events\right)$,
then the norm of their difference is $\left\Vert \rho-\rho'\right\Vert \le\delta$.\end{thm}

Given a state~$\rho'$ consistent with $\iota\circ\mu_{\rho}^{\mathrm{B}}$,
we have $\mu_{\rho'}^{\mathrm{B}}\left(\proj{\psi}\right)\in\iota\left(\mu_{\rho}^{\mathrm{B}}\left(\proj{\psi}\right)\right)$
for any 1-dimensional projector~$\proj{\psi}$. Since the maximum
length of the intervals in $\mathscr{I}$ is $\delta$, it is also
the upper bound of the difference $\left|\mu_{\rho'}^{\mathrm{B}}\left(\proj{\psi}\right)-\mu_{\rho}^{\mathrm{B}}\left(\proj{\psi}\right)\right|=\left|\melem{\psi}{\rho-\rho'}{\psi}\right|$.
Since $\rho-\rho'$ is Hermitian, $\max_{\ip{\psi}{\psi}=1}\left|\melem{\psi}{\rho-\rho'}{\psi}\right|$is
the maximum absolute value of the eigenvalues of $\rho-\rho'$~\citep{544199},
and equal to $\left\Vert \rho-\rho'\right\Vert $~\citep{GolubVanLoan1996,Foucart2012}.
This implies $\left\Vert \rho-\rho'\right\Vert \le\delta$ by definition.
